---
title: "Market basket Analysis"
author: "Soheil Garfami"
date: "2024-08-19"
output: 
  html_document: 
    toc: true
---
# **Introduction**

Hi! In this kernel we are going to use the **Apriori algorithm** to perform a **Market Basket Analysis**. It's a technique used by large retailers to uncover associations between items.
It works by looking for combinations of items that occur together frequently in transactions, providing information to understand the purchase behavior. The outcome of this type of technique is, 
in simple terms, a set of **rules** that can be understood as **“if this, then that”**.

### Key Objectives

***Market Basket Analysis &  Association Rules***
   
In this section we use association rules and the a priori algorithm to identify purchase patterns and answer the following questions: 

* Do customers purchase items together frequently and which products are most often purchased together ?

* Can we use this information to recommend other products based on a customer’s cart ?
    
    
### Importing the data

First, let's import the required libraries.
```{r message=FALSE, warning=FALSE}
library(arules)
library(arulesViz)
library(tidyverse)
library(ggplot2)
library(DataExplorer)
library(gridExtra)
```

let's see the data set that we built by joining all datasets
```{r}
JOINED <- read.csv("data/JOINED.csv")
head(JOINED)
```
   
To apply the `Apriori` algorithm in `market basket analysis`, we need to have a **transactions** dataset. We can import the dataset in the required transaction format using the `arules` library. For this, we'll need the columns for transaction ID and product name.


```{r}
JOINED[1,c(1,11)]
```
As you can see, the `transactionId` and `productName` columns are in positions 1 and 11, respectively.

let's imprort the dataset as transactions. (columns 1 and 11)
```{r}
rm(JOINED)
trans <- read.transactions("data/JOINED.csv" , format = "single" , sep = "," , cols = c(1 ,11))
trans
```
let's see the first transaction 

```{r}
inspect(trans[1])
```

With the `itemFrequencyPlot`, we can again visualize a graph of the most popular products.

    
```{r}
itemFrequencyPlot(trans ,topN=10 , type="absolute" ,col="lightcyan2",xlab="Item name", 
                  ylab="Frequency (absolute)", main="Absolute Item Frequency Plot")
```

<hr>

    
# **Selecting the Optimal Values for Parameters**

    
The Apriori algorithm generates association rules for a given dataset. An association rule suggests that if item A occurs, then item B is likely to occur with a certain probability. 

To extract these rules using the Apriori algorithm, there are three key arguments that are crucial for obtaining the best rules:

- **Support**: Indicates how frequently the itemset appears in the dataset.
- **Confidence**: Reflects how often the rule has been found to be true.
- **Lift**: Represents the ratio of the observed support to that expected if X and Y were independent. If the lift is greater than 1, it indicates the strength of the relationship between the two items, showing how dependent they are on each other.

Now, we will use the `gridExtra` package to build different rules with varying parameters and plot them to determine which set of rules is the most suitable.

```{r}
# Support and confidence values
supportLevels <- c( 0.05, 0.01, 0.005 , 0.001)
confidenceLevels <- c(  0.7, 0.6, 0.5, 0.4, 0.3, 0.2, 0.1 , 0.05 , 0.1)

# Empty integers 
rules_sup5 <- integer(length=9)
rules_sup1 <- integer(length=9)
rules_sup0.5 <- integer(length=9)
rules_sup0.1 <- integer(length=9)
```

We will test `support values` of 5, 1, 0.5, and 0.1, along with `confidence values` of 0.7, 0.6, 0.5, 0.4, 0.3, 0.2, 0.1, 0.05, and 0.01 to determine which combination of these two parameters works best together.

```{r echo=TRUE, message=FALSE, warning=FALSE}

# Apriori algorithm with a support level of 5%
for (i in 1:length(confidenceLevels)) {
  
  rules_sup5[i] <- length(apriori(trans, parameter=list(sup=supportLevels[1], 
                                                         conf=confidenceLevels[i],target="rules"), 
                                control = list(verbose = FALSE)))
}

# Apriori algorithm with a support level of 1%
for (i in 1:length(confidenceLevels)){
  
  rules_sup1[i] <- length(apriori(trans, parameter=list(sup=supportLevels[2], 
                                                        conf=confidenceLevels[i], target="rules"), 
                                control = list(verbose = FALSE)))
}

# Apriori algorithm with a support level of 0.5%
for (i in 1:length(confidenceLevels)){
  
  rules_sup0.5[i] <- length(apriori(trans, parameter=list(sup=supportLevels[3], 
                                                        conf=confidenceLevels[i],target="rules"), 
                                control = list(verbose = FALSE)))
}

# Apriori algorithm with a support level of 0.1%
for (i in 1:length(confidenceLevels)){
  
  rules_sup0.1[i] <- length(apriori(trans, parameter=list(sup=supportLevels[4], 
                                                          conf=confidenceLevels[i], target="rules"), 
                                control = list(verbose = FALSE)))
}

```

```{r}
rules_sup5 
rules_sup1 
rules_sup0.5 
rules_sup0.1
```

***Now, let's plot the number of rules generated for each confidence level at each support level.***

```{r warning=FALSE}

# Number of rules found with a support level of 5%
plot1 <- qplot(confidenceLevels, rules_sup5, geom=c("point", "line"), 
               xlab="Confidence level", ylab="Number of rules found", 
               main="Apriori with a support level of 5%") +
  theme_bw()

# Number of rules found with a support level of 1%
plot2 <- qplot(confidenceLevels, rules_sup1, geom=c("point", "line"), 
               xlab="Confidence level", ylab="Number of rules found", 
               main="Apriori with a support level of 1%") + 
  scale_y_continuous(breaks=seq(0, 10, 2)) +
  theme_bw()

# Number of rules found with a support level of 0.5%
plot3 <- qplot(confidenceLevels, rules_sup0.5, geom=c("point", "line"), 
               xlab="Confidence level", ylab="Number of rules found", 
               main="Apriori with a support level of 0.5%") + 
  scale_y_continuous(breaks=seq(0, 50, 10)) +
  theme_bw()

# Number of rules found with a support level of 0.1%
plot4 <- qplot(confidenceLevels, rules_sup0.1, geom=c("point", "line"), 
               xlab="Confidence level", ylab="Number of rules found", 
               main="Apriori with a support level of 0.1%") + 
  scale_y_continuous(breaks=seq(0, 130, 20)) +
  theme_bw()

# Subplot
grid.arrange(plot1, plot2, plot3, plot4, ncol=2)

```

<hr>


***Let's plot them all together so we can compare them.***
    
    
```{r}
num_rules <- data.frame(rules_sup5, rules_sup1, rules_sup0.5, rules_sup0.1, confidenceLevels)

# Number of rules found with a support level of 10%, 5%, 1% and 0.5%
ggplot(data=num_rules, aes(x=confidenceLevels)) +
  
  # Plot line and points (support level of 10%)
  geom_line(aes(y=rules_sup5, colour="Support level of 5%")) + 
  geom_point(aes(y=rules_sup5, colour="Support level of 5%")) +
  
  # Plot line and points (support level of 5%)
  geom_line(aes(y=rules_sup1, colour="Support level of 1%")) +
  geom_point(aes(y=rules_sup1, colour="Support level of 1%")) +
  
  # Plot line and points (support level of 1%)
  geom_line(aes(y=rules_sup0.5, colour="Support level of 0.5%")) + 
  geom_point(aes(y=rules_sup0.5, colour="Support level of 0.5%")) +
  
  # Plot line and points (support level of 0.5%)
  geom_line(aes(y=rules_sup0.1, colour="Support level of 0.1%")) +
  geom_point(aes(y=rules_sup0.1, colour="Support level of 0.1%")) +
  
  # Labs and theme
  labs(x="Confidence levels", y="Number of rules found", 
       title="Apriori algorithm with different support levels") +
  theme_bw() +
  theme(legend.title=element_blank())
```

<hr>


# **Analysis & Association Rules**

Let's use a support level of 1% and a confidence level of 0% for this analysis.
`Note: In the real world, this level of support and confidence may not be sufficient, and we typically rely on stronger evidence`.

```{r}
rules <- apriori(trans, parameter=list(sup=supportLevels[4], 
                                                   conf=confidenceLevels[3], target="rules"), 
                                control = list(verbose = FALSE))
rules
inspect(head(rules))
```
    
```{r}
plot(rules, measure=c("support", "lift"), shading="confidence")
```
    
```{r}
plot(rules, method="graph")
```


<hr>

***Now, I want to explore 4-product associations to determine if purchasing products A, B, and C increases the likelihood of purchasing product D.***

```{r}
rules <- apriori(trans, parameter=list(sup=supportLevels[4], 
                                                   conf=confidenceLevels[4],minlen =4 , maxlen=4, target="rules"), 
                                control = list(verbose = FALSE))
rules
```


  

```{r}
inspect(head(rules))
```
<hr>

# **Conclusion**

* Using association rules we have identfied items that customers purchase together frequently and which products are most often purchased together

* In addition, we have an apriori model that we can use to recommend instcart customer products based their existing cart !
